---
title: 'mrIML: Multivariate (multi-response) interpretable machine learning'

output: github_document

---

<img src = "https://user-images.githubusercontent.com/33707823/88988817-531ad080-d31d-11ea-8d76-f1ad0506e405.png" width="200" height="200"/>

This package aims to enable users to build and interpret multivariate machine learning models harnessing the tidyverse (tidy model syntax in particular). This package builds off ideas from Gradient Forests (Ellis et al 2012), ecological genomic approaches (Fitzpatrick and Keller, 2014) and multi-response stacking algorithms (Xing et al 2019).

This package can be of use for any multi-response machine learning problem, but was designed to handle data common to community ecology (site by species data) and ecological genomics (individual or population by SNP loci).

## Installation 

Install the stable version of the package:

```{r, eval = FALSE}
#install.packages("devtools")
devtools::install_github('nfj1380/mrIML')
```

## Quick start

`{mrIML}` is designed to be used with a single function call or to be used in an ad-hoc fashion via individual function calls. In the following section we give an overview of the simple use case. For more on using each function see the [function documentation](xx). The core functions for both regression and classification are: [`mrIMLpredicts`](xx),
[`mrIMLperformance`](xx), and 
[`mrInteractions`](xx),for plotting and visualization
[`mrVip`](xx), [`mrFlashlight`](xx), and[`plot_vi`](xx). Estimating the interactions
alone can be substantially computationally demanding depending on the number
of outcomes you want to test. The first step to using the package is to load it as follows.

```{r, message = FALSE}
library(mrIML)
```



## References
Xing, L, Lesperance, ML and Zhang, X (2020). Simultaneous prediction of multiple outcomes using revised stacking algorithms. Bioinformatics, 36, 65-72.

Fitzpatrick, M.C. & Keller, S.R. (2015) Ecological genomics meets community-level modelling of biodiversity: mapping the genomic landscape of current and future environmental adaptation. Ecology Letters 18, 1–16.

Ellis, N., Smith, S.J. and Pitcher, C.R. (2012), Gradient forests: calculating importance gradients on physical predictors. Ecology, 93: 156-168. doi:10.1890/11-0252.1
