---
title: "Microbiome data curation"
opengraph:
  image:
    src: 
output:
  rmarkdown::html_vignette: default
  github_document: default
  css: extra.css
vignette: >
  %\VignetteIndexEntry{Microbiome data curation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
    fig.width = 8, fig.height = 6, fig.align = "center",
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```
In this vignette, we will walk you through how to curate your microbiome data extracted from pipelines such as QIIME 2 into a format suitable for MrIML. The main problem with using the raw data as the ASV names a long character vectors of nucleotide data which will make interpretation impossible. The data set is just 10 ASVs from the ostrich case study in Fountain-Jones et al (2024) to keep things simple. We will check out the data and load the required packages.

```{r include=FALSE}
pacman::p_load('MRFcov', 'mrIML', 'tidyverse', 'future.apply','tidymodels', 'finetune',
               'themis', 'vip', 'flashlight', 'iml', 'vivid', 'igraph', 'ggnetwork', 'network',
               'gridExtra', 'xgboost', 'brulee', 'fastshap', 'tabnet', 'bonsai',
               'parsnip','data.table', 'cowplot', 'progress', 'hstats', 'geosphere')


#raw ASV data
asv_data <- otu_table_ostrich
str(asv_data)

#corresponding taxa table
#load("~/MrIML/mrIML/data/asv_taxa_table.rda")
asv_taxa_table <- taxa_table_ostrich
str(asv_taxa_table)
```

##Preparing the ASV table

The first step is to prepare the taxa table by removing columns and making sure columns aren't identical. This will help when we start to create meaningful names for the ASVs.

```{r cars}
#----------------------------------------------------
# Prepare ASV table
#----------------------------------------------------

#ASV column not needed
asv_taxa_table$ASV <- NULL
asv_taxa_table$Kingdom<- NULL
asv_taxa_table$Class<- NULL
asv_taxa_table$Phylum<- NULL

#create_names
create_name <- function(row) {
  non_na_values <- na.omit(row)
  unique_values <- unique(non_na_values)  # Remove duplicates
  
  if (length(unique_values) > 3) {
    unique_values <- unique_values[1:2]
  }
  
  # Extract genus and species names if available
  family <- ifelse("Family" %in% names(row), row["Family"], NA)
  genus <- ifelse("Genus" %in% names(row), row["Genus"], NA)
  species <- ifelse("Species" %in% names(row), row["Species"], NA)
  
  # Combine genus, species, and unique values into a name
  name_parts <- c(family,genus, species, unique_values)
  name_parts <- name_parts[!is.na(name_parts)]  # Remove NA values
  return(paste(name_parts, collapse = "_"))
}

# Create the 'Name' column
asv_taxa_table$Name <- apply(asv_taxa_table, 1, create_name)

```

## Creating names

The next functions take the ASV table data and make unique names for the new column in the taxa table.

```{r,, echo=FALSE}
#make names unique
make_names_unique <- function(names_vector, remove_duplicates = character(0)) {
  # Remove specified strings (e.g., "Clostridiales") from the names vector
  names_vector <- names_vector[!names_vector %in% remove_duplicates]
  
  # Create a named vector to keep track of counts for each name
  name_counts <- numeric(length(names_vector))
  names(name_counts) <- names_vector
  
  # Initialize a result vector
  result_names <- character(length(names_vector))
  
  for (i in seq_along(names_vector)) {
    name <- names_vector[i]
    
    # Check if the name exists in name_counts
    if (!is.na(name_counts[name])) {
      # Check if the name has occurred before
      if (name_counts[name] > 1) {
        # Append a number to make the name unique
        result_names[i] <- paste0(name, "_", name_counts[name])
      } else {
        result_names[i] <- name
      }
      
      # Update the count for this name
      name_counts[name] <- name_counts[name] + 1
    } else {
      # If name doesn't exist in name_counts, treat it as the first occurrence
      result_names[i] <- name
      name_counts[name] <- 1
    }
  }
  
  return(result_names)
}


asv_taxa_table$Name_f <- make_names_unique(asv_taxa_table$Name)

replace_empty_names <- function(names_vector) {
  # Find indices where names are empty
  empty_indices <- names_vector == ""
  
  # Count occurrences of empty names
  empty_counts <- cumsum(empty_indices)
  
  # Replace empty names with "Taxon" followed by count
  names_vector[empty_indices] <- paste0("Taxon ", empty_counts[empty_indices])
  
  return(names_vector)
}

asv_taxa_table$Name_comp <-replace_empty_names(asv_taxa_table$Name_f)

make_names_unique3 <- function(names_vector) {
  # Initialize a counter to keep track of occurrences of each name
  name_counts <- list()
  
  # Initialize a result vector to store the modified names
  result_names <- character(length(names_vector))
  
  for (i in seq_along(names_vector)) {
    name <- names_vector[i]
    
    # Check if the name is empty or has occurred before
    if (name == "" || is.null(name_counts[[name]]) || name_counts[[name]] > 0) {
      # Increment the count for this name
      count <- ifelse(is.null(name_counts[[name]]), 1, name_counts[[name]] + 1)
      name_counts[[name]] <- count
      
      # Append a number to make the name unique
      result_names[i] <- paste0(name, ".", count)
    } else {
      result_names[i] <- name
      # First occurrence of this name, initialize count to 1
      name_counts[[name]] <- 1
    }
  }
  
  return(result_names)
}

asv_taxa_table$Name_comp1 <- make_names_unique3(asv_taxa_table$Name_comp)
```


Now we can start fomatting our Y (response data) ready for MrIML. This includes filtering rare and common taxa.

```{r}

#make sure names match
final_ASV_table <- asv_data
final_ASV_table$ASV <- asv_taxa_table$Name_comp1 

#add row names
final_ASV_table_r <- final_ASV_table %>% 
  column_to_rownames(var='ASV')
#get into right format for MrIML
final_ASV_table_df <-as.data.frame(t(final_ASV_table_r ))

#make presence/absence
pa_ASV_table <- final_ASV_table_df %>% 
  mutate_all(~ ifelse(. > 0, 1, .))

#remove rare and common ASVs
Y <- filterRareCommon(pa_ASV_table , lower=0.1, higher=0.9) %>% 
  dplyr::select(sort(names(.))) #0.2

colnames(Y) <- sub(".*_(.*_.*)$", "\\1", colnames(Y))

## Find duplicated column names
duplicated_cols <- duplicated(colnames(Y))

colnames(Y) <- make.names(colnames(Y), unique = TRUE)

#more tidying
Y <- Y %>% 
  rename_all(~make.names(str_remove_all(., "`")))

# Shorten the part to the left of the underscore to 4 characters

abbreviate_names <- function(names_vector) {
  abbreviated_names <- sapply(strsplit(names_vector, "_"), function(parts) {
    first_part <- substr(parts[1], 1, 4)
    second_part <- parts[2]
    return(paste(first_part, second_part, sep = "_"))
  })
  return(abbreviated_names)
}

# Use the function to shorten the names(
modified_col_names <- abbreviate_names(colnames(Y))

colnames(Y) <- modified_col_names
glimpse(Y)



```

Now we are good to go and the MrIML 2.0 grahical network models can be generated!


